Discriminating causes \\from multivariate observations:\\the back-to-back regression

### 2020-03-04 paper/paper.tex
hard rewrap, causal terminology

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{iclr2020_conference,times,wrapfig}

@@

\title{Measuring causal influence with\\ back-to-back regression: the linear case}

\author{%
  Jean-Remi King\\
  CNRS\\
  \texttt{jeanremi.king@gmail.com} \\
  \And
  Fran\c{c}ois Charton\\
  Facebook AI\\
  \texttt{fcharton@fb.com}\\
  \And
  David Lopez-Paz\\
  Facebook AI\\
  \texttt{dlp@fb.com}
  \And
  Maxime Oquab\\
  Facebook AI\\
  \texttt{qas@fb.com}
}

@@

\begin{abstract}
Identifying causes from observations can be particularly challenging when i) potential factors are difficult to manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
method designed to efficiently measure, from a set of co-varying factors, the causal
influences that most plausibly account for multidimensional observations. After
proving the consistency of B2B and its links to other linear approaches, we show that our method outperforms
least-squares regression and cross-decomposition techniques (e.g.
canonical correlation analysis and partial least squares) on causal
identification. Finally, we apply B2B to
neuroimaging recordings of 102 subjects reading word sequences. The results
show that the early and late brain representations, caused by low- and high-level
word features respectively, are more reliably detected with B2B than with other standard techniques.

@@

\begin{figure}[t!]
    \centering
    \begin{tikzpicture}
    \newcommand\posY{0}
    \newcommand\posX{3}
    \newcommand\posE{5}
    \newcommand\posN{7}
    \newcommand\posF{10}

    \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
    \node[] (eq) at (1.5, 0){$=$};
    \node[] (times) at (4, 0){$\times$};
    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
    \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};

    \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
    \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
    \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
    \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
    \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
    \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
    \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
    \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
    \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
    \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);

    \node[] (plus) at (6, 0){$+$};
    \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=blue!10] (N) at (\posN, 0){};
    \node[thick, draw=black, minimum height=1cm, minimum width=2cm, fill=red!10] (F) at (\posF, 0){};
    \draw [thick] (2.25, -1.5) to [round left paren ] (2.25, 1.5);
    \draw [thick] (7.75, -1.5) to [round right paren ] (7.75, 1.5);

    \node[] (times2) at (8.5, 0){$\times$};

    \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
    \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
    \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d_x \times d_x}$}};
    \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
    \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};

    \node[] (labY) at (\posY, 2){Observations};
    \node[] (labX) at (\posX, 2){Factors};
    \node[] (labE) at (\posE, 2){Cause};
    \node[] (labE) at (\posE, 1.7){selection};
    \node[] (labN) at (\posN, 2){Noise};
    \node[] (labF) at (\posF, 2){Cause-effect};
    \node[] (labF) at (\posF, 1.7){mapping};

    \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
    \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
    \node[] (reg1) at (5.5,-3.25) {$\hat{E} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
    \end{tikzpicture}
    \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
    \label{fig:b2b}
\end{figure}

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{breiman1996bagging}.
%
To accelerate this ensembling procedure, we implemented an efficient leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes} % http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf
, as follows:
%

>>>>

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{times,wrapfig}
\usepackage{authblk}
\usepackage{apacite}
\usepackage{natbib}
\usepackage{biblatex}

@@

\title{Discriminating causes \\from multivariate observations:\\the back-to-back regression}

\author[1, 2]{Jean-Rémi King}
\author[2]{Fran\c{c}ois Charton}
\author[2]{David Lopez-Paz}
\author[2]{Maxime Oquab}
\affil[1]{Laboratoire des systèmes perceptifs, PSL University, CNRS}
\affil[2]{Facebook AI}
\date{}

% \author{%
%   Jean-Rémi King\\
%   CNRS\\
%   \texttt{jeanremi.king@gmail.com} \\
%   \And
%   Fran\c{c}ois Charton\\
%   Facebook AI\\
%   \texttt{fcharton@fb.com}\\
%   \And
%   David Lopez-Paz\\
%   Facebook AI\\
%   \texttt{dlp@fb.com}
%   \And
%   Maxime Oquab\\
%   Facebook AI\\
%   \texttt{qas@fb.com}
% }

@@

\begin{abstract}
Identifying causes solely from observations can be particularly challenging when
i) potential factors are difficult to manipulate individually and ii)
observations are complex and multi-dimensional. To address this issue, we
introduce ``Back-to-Back'' regression (B2B), a method designed to efficiently
measure, from a set of colinear factors, those that most plausibly account for
multidimensional observations. After proving the consistency of B2B and its
links to other linear approaches, we show that our method outperforms
least-squares regression and cross-decomposition techniques (e.g. canonical
correlation analysis and partial least squares) on causal identification.
Finally, we apply B2B to neuroimaging recordings of 102 subjects reading word
sequences. The results show that the early and late brain representations,
caused by low- and high-level word features respectively, are more reliably
detected with B2B than with other standard techniques.

@@

\include{fig_problem.tex}

@@

Performing two regressions over the same data sample can result in overfitting,
as spurious correlations in the data absorbed by the first regression will be
leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$
and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second
regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2B is
repeated over many random splits, and the final estimate $\hat E$ of the causal
influence matrix is the average over the estimates associated to each split
\citep{breiman1996bagging}.
%
To accelerate this ensembling procedure, we implemented an efficient
leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes} %
http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf, as follows:
%
